Incorporating periodic variability in hidden Markov models for animal movement

### 2016-02-01 paper3.Rnw
tweaking paper, working on figs

\documentclass{article}
\usepackage{natbib}
\usepackage{amsmath}
\usepackage[utf8]{inputenc} % for accented characters
%% stuff for editing
%\usepackage[markup=nocolor,addedmarkup=bf,deletedmarkup=sout]{changes}
%% to suppress notes & comments: \usepackage[final]{changes}
\usepackage[backgroundcolor=lightgray,textsize=tiny]{todonotes}
\usepackage{setspace}
\bibliographystyle{chicago}
\title{Incorporating periodic variability in hidden Markov models for animal movement}
\author{Michael Li and Benjamin M. Bolker }
\date{\today}

@@

\doublespacing
<<pkgs,message=FALSE,warning=FALSE,echo=FALSE>>=
library("ggplot2")
library("depmixS4")
library("dplyr")
library("reshape2")
library("plyr")
@

@@

Our primary goal in this paper is to introduce the use of
HMMs with temporally varying transition probabilities -- in particular,
transition probabilities that follow a diurnal cycle -- for modeling
animal movement recorded over long time scales.  
In addition to simulation-based examples,
we also re-analyze data from 
\cite{kerk2015hidden}, who used temporally homogeneous hidden semi-Markov
models (HSMMs: an extension of HMMs that allow
flexible modelling of the distribution of \emph{dwell times}, the 
lengths of consecutive occupancy of a behavioural state) to
describe the movement and putative underlying
behavioural states of Florida panthers.

@@

GPS collars were fitted to 18 Florida panthers in 2005-2012 by Florida
Fish and Wildlife and Conservation Commission staff using
trained hounds and houndsmen. Of these animals, 13 had 
sufficient data to be used by \cite{kerk2015hidden}.
\todo{not sure quite where we should use {\tt citeauthor} vs. {\tt cite} \ldots}
Here we focus
on the four cats with the most data, in part because our goal is to understand
the issues that arise when simple models are fitted to large data sets, and 
in part because the general trend in telemetry studies is toward larger data sets. \todo{give cat IDs - these are shown in Fig S2a of vdK.  Give threshold amount of data; justify/describe how we decided.  Figures seem to focus on cat 130 - do we want to include some more info (maybe in suppl) about other cats? ML: Yes, I have all the plots ready in the figures folder. (Note, the public catID is different from what Madelon used, so we have to be careful)}
As is typical in studies of animal movement, 
we took first differences of the data by decomposing
contiguous sequences of hourly GPS coordinates into 
successive step lengths (in meters) and turning angles (in radians)
\citep{turchin1998quantitative,kerk2015hidden}.

\begin{table}[]
\centering
\caption{Cat-ID}
\label{Cat-ID}
\begin{tabular}{ccc}
 (van de Kerk 2015) & (IR@UF)  & Number of Obserations  \\
  130 & 1 & 10286    \\
  131 & 2  & 9458    \\
 48& 14  & 14645    \\
 94& 15  & 10250  
\end{tabular}
\end{table}

\cite{kerk2015hidden} used hidden semi-Markov models (HSMM), 
an extension of HMM that permits explicit modelling of dwell times 
\citep{langrock_flexible_2012}, considering both Poisson and
negative binomial distributions for dwell times.  As shown 
\todo{I was going to say ``but not emphasized'', but that seems snarky}
by \cite{kerk2015hidden} (Figure S3b, top row, middle panel),
the estimated shape parameter of the negative binomial dwell time
distribution was typically close to 1 ($\approx 0.4-1.6$; confidence
intervals were not calculated), implying that a geometric distribution
(i.e., negative binomial with shape=1) might be adequate. In turn,
this suggests that we might not lose much accuracy by reverting to 
a simpler HMM framework, which makes precisely this assumption.
\todo{could say here (or maybe better in the conclusions) that
revisiting HSMM in this context would be worthwhile future work. 
ML: conclusion is good, I still need to add viterbi paragraph}

\cite{kerk2015hidden} considered time-homogeneous models with 
a variety of candidate distributions %
--- log-Normal, Gamma, and Weibull distributions for step lengths
and von Mises and wrapped Cauchy distributions for the turning angle --- %
concluding on the basis of the Akaike information criterion (AIC) 
that Weibull step length and wrapped Cauchy turning angle distributions
were best.

Our analysis again aims for simplicity and qualitative
conclusions rather than for picking the very best
predictive model; we focus on models that treat each step as
a univariate, log-Normally distributed observation, glossing
over both the relatively subtle distinctions in shape between
the three candidate step-length distributions and the effects
of considering multivariate (step length plus turning angle)
observations. (Note that most movement analyses, including
\cite{kerk2015hidden}, are only partially multivariate, 
treating step length and turning angle at a particular time
as independent observations while neglecting possible correlations
between them.)  However, we do briefly compare all three
step-length distributions, with and without incorporating
a von Mises-distributed turning angle in the model (Figure~1).

\cite{kerk2015hidden} used the Bayesian (Schwarz) information criterion (BIC)
to test the relative penalized goodness of fit for
models ranging from 2 to 6 latent states.
\todo{not sure about terminology here (should be consistent about
what we call the modes/latent states/whatever) ML: latent states}
In general, BIC values decreased as the number of states
increased from 
3 to 6 states,
\todo{should decide on rules for numbers vs numerals (3 vs three) and be consistent. ML: numeric}
suggesting that the 6-state model was
favoured statistically; however,
the authors used 3-state models in most
of their analyses for ease of biological interpretation.
We follow \cite{kerk2015hidden}'s lead in using BIC-optimality (i.e., minimum
BIC across a family of models) as the criterion for identifying
the best model, because we are interested in explaining the 
data generation process by identifying the `true' number 
of underlying movement states.  Using BIC also simplifies 
evaluation of model selection procedures; 
it is easier to test whether our model selection
procedure has selected the model used to simulate the data,
rather than testing whether it has selected the model with
the minimal Kullback-Leibler distance
\citep{Richards2005}. We recognize that 
ecologists will often be interested in maximizing predictive
accuracy rather than selecting a true model, and that as
usual in ecological systems the true model will be far more
complicated than any candidate model \citep{BurnhamAnderson1998};
we suspect that the qualitative conclusions stated here
for BIC-optimality will carry over to analyses using AIC instead.

\todo{The following description of the V. algorithm is not coherent; need to give
just a \emph{brief} definition here (since we are going to use it) and criticize/
explain shortcomings later. ML: I moved the whole previous paragraph to the end where we actually talk about viterbi}

@@

Eq.~\ref{eq:HMMlik} gives the likelihood of the observed sequence 
given (conditional on) a particular
hidden sequence. 
In order to calculate the overall, unconditional (or marginal) 
likelihood of the 
observed sequence, we need to average over all possible hidden sequences. 
While naive approaches to this averaging process there are several
more efficient algorithms for computing the marginal likelihood and
numerically estimating parameters \citep{zucchini_hidden_2009};
we use the implementations in the {\tt depmixS4} package for R
\citep{visser2010depmixs4,R}.

\todo{where do we actually focus on these three states?  Think this could probably be introduced in the Discussion, if we need it. ML: I took it out and gave a half a line in model evaluation subsection}

For any $n$-state HMM, we need to define a $n \times n$ matrix that specifies the probabilities $\pi_{ij}$ of being in movement states $j$ at time $t+1$ given that the individual is in state $i$.  However, this matrix needs to respect the constraints that (1) all probabilities are between 0 and 1 and (2) transition probabilities out of a given state sum to 1. The FMM is a special case of HMM where; it can be modelled in the HMM framework by setting the transition probabilities:

\begin{equation}
\begin{split}
\pi_{ij} = \pi_{i*j},     \forall i 
\end{split}
\end{equation}

Therefore, as is standard for HMMs with covariates \citep{visser2010depmixs4}, we define this multinomial logistic model in terms of a linear predictor $\eta_{ij}$, where $\eta_{i1}$ is unnecessary (i.e. we have only $n \times (n-1)$ distinct parameters; we nevertheless index $j$ from 2 to $n$ for notational clarity):
\begin{equation}
\begin{split}
\pi_{ij} & = \exp(\eta_{ij}(t))/\left(1+\sum_{j=2}^{n}\exp(\eta_{ij}(t))\right), \textrm{ for } j={2,...,n} \\
\pi_{i1} & = 1 - \sum_{j=2}^{n}\pi_{ij}
\end{split}
\end{equation}

We considered three different transitions models for diurnal variation in behaviour; extension of ``Switch with covariates" by \cite{morales_extracting_2004} to account for temporal hetergeneitiy \todo{second clause isn't clear; expand? ML: okay now? }.

\begin{description}
\item[Multiple block transition] Here we assume piecewise-constant transition probabilities. The transition probability $\pi_{ij}$ is a function of time (hour of day), where it is assigned to one of $M$ different time blocks:
$$
\eta_{ij}(t) = \sum_{m=1}^{M}a_{ijm} \delta_{m=t}
$$
where $a_{ijm}$ are parameters, and $\delta_{m=t}$ is a Kronecker delta or  indicator variable (1 for the time block at the corresponding time $t$, and 0 otherwise).  
\item[Quadratic transition model] We assume the elements of the linear predictor
are quadratic functions of hour, with an additional diurnal continuity constraint that $\eta_{ij}(0)=\eta_{ij}(24)$ \todo{am I right/do we actually impose this constraint? give a bit more detail ...}
$$
\eta_{ij}(t) = b_{ij1}+b_{ij2}(\frac{t}{24})+b_{ij3}(\frac{t}{24})^{2}
$$
\item[Sinusoidal transition model] A sinusoidal model with a period of 24 hours is identical in complexity to the quadratic model, \todo{apparently true from eqs here, but not true if we constrain the quadratic model??} but automatically satisfies the diurnal continuity constraint.
$$
\eta_{ij}(t)= b_{ij1}+b_{ij2}\cos\left(\frac{2\pi t}{24}\right)+b_{ij3}\sin\left(\frac{2\pi t}{24}\right)
$$
\end{description}

Model complexity and number of parameters increase as the number of latent states increase, FMM to HMM, and lastly, FMM and HMM incorportating temportal heterogeneity.

\begin{table}[]
\centering
\caption{Number of Parameters for each model type}
\label{Cat-ID}
\begin{tabular}{cccc}
 Type & Model  & Number of States & Number of Parameters  \\
  FMM & FMM & 3 & 9    \\
  FMM & FMM & 4 & 12    \\
  FMM & FMM & 5 & 15   \\
  FMM & FMM & 6 & 18  \\
  FMM + TH & FMM + TH (Sin)  & 3 & 15    \\
  FMM + TH & FMM + TH (Sin)  & 4 & 20    \\
  FMM + TH & FMM + TH (Sin)  & 5 & 25    \\
  FMM + TH & FMM + TH (Sin)  & 6 & 30    \\
  HMM & HMM & 3 & 18    \\
  HMM & HMM & 4 & 28    \\
  HMM & HMM & 5 & 40   \\
  HMM & HMM & 6 & 54  \\
  HMM + TH & HMM + TH (Block)  & 3 & 36    \\
  HMM + TH & HMM + TH (Block)  & 4 & 60    \\
  HMM + TH & HMM + TH (Block)  & 5 & 90    \\
  HMM + TH & HMM + TH (Block)  & 6 & 126    \\
  HMM + TH & HMM + TH (Quad)  & 3 & 36    \\
  HMM + TH & HMM + TH (Quad)  & 4 & 60    \\
  HMM + TH & HMM + TH (Quad)  & 5 & 90    \\
  HMM + TH & HMM + TH (Quad)  & 6 & 126    \\
  HMM + TH & HMM + TH (Sin)  & 3 & 36    \\
  HMM + TH & HMM + TH (Sin)  & 4 & 60    \\
  HMM + TH & HMM + TH (Sin)  & 5 & 90    \\
  HMM + TH & HMM + TH (Sin)  & 6 & 126    \\
  HMM + TH & HMM + TH (Hourly)  & 3 & 225    \\
  HMM + TH & HMM + TH (Hourly)  & 4 & 396

\end{tabular}
\end{table}

\todo{Discuss model complexity/parameter-number computations here, or with each model.  Also discuss the ``hourly'' model here.}

@@

\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}
\begin{figure}
<<All mod BIC plot, warning=FALSE, echo=FALSE, message=FALSE>>=
load("cat1allmod.RData")
ggplot(allmodsumdf, aes(x=nstates,y=deltaBIC,colour=model))+facet_wrap(~ model,ncol=4)+
    scale_y_continuous()+
    scale_size_continuous( name="Num of Parameters") +
    #    geom_point(aes(size=parameters))+
    labs(y=expression(paste(Delta, "BIC")),
         x="Number of States")  +
    geom_line(aes(colour=model),size=0.5) +theme_bw() +
    ggtitle("Adjusted BIC Model Comparisons")

@
\caption{Model Reduction Process via Individual based Model Adjusted BIC }
\label{fig:BICred_plot}
\end{figure}
\end{knitrout}

@@

\begin{knitrout}
\begin{figure}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}

<<Adjusted BIC comparisons, warning=FALSE, echo=FALSE, message=FALSE>>=
load("cat1sumdat.RData")
ggplot(sumdat, aes(x=nstates,y=deltaBIC,colour=model))+facet_wrap(~ type,ncol=4)+
    scale_y_continuous()+
    scale_size_continuous( name="Num of Parameters") +
    geom_point(aes(size=parameters))+
    labs(y=expression(paste(Delta, "BIC")),
         x="Number of States") +
    geom_line(aes(colour=model),size=0.5) +theme_bw() +
    ggtitle("Adjusted BIC Model Comparisons")

\caption{Overall Adjusted BIC Comparisons between different HMM models}
\end{figure}
\end{knitrout}

@@

\begin{knitrout}
\begin{figure}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}

<<avg step length by time, message=FALSE, echo=FALSE, warning=FALSE>>=
  simdf <- simdat %>% dplyr:::select(-obs) %>% group_by(time) %>% summarise_each(funs(mean))
  obsdf <- simdat %>% dplyr:::select(c(obs,time)) %>% filter(!is.na(obs)) %>% group_by(time) %>% summarise_each(funs(mean)) %>% dplyr:::select(obs)
  avgdf <- cbind(simdf,obsdf)
  avgdf2 <- melt(avgdf,'time')
ggplot(avgdf2,aes(x=time,y=value,colour=variable)) + 
    geom_point() +geom_line()+ xlab("Time of Day") + ylab("Step lengths (log10)")+
    scale_x_continuous(breaks=0:23) + theme_bw() + 
    ggtitle("Average Step Lengths at Different Hours of the Day")

\caption{Average step length by time of day}
\end{figure}

\end{knitrout}

@@

\begin{knitrout}
\begin{figure}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}
<<acf plot, message=FALSE, echo=FALSE, warning=FALSE>>=
acffun <- function(sim){
  a <- acf(sim,na.action=na.pass,plot=FALSE)
  df <- data.frame(ACF=c(a$acf),lag=a$lag)
  return(df)
}

  temp <- simdat %>% dplyr:::select(-time)%>%ldply(.,acffun)
  ggplot(temp,aes(lag,ACF,colour=.id))+
    geom_point()+geom_line() +theme_bw() +
    ggtitle("ACF plot of BIC Selected Models")

@
\caption{ACF }
\end{figure}
\end{knitrout}

@@

\begin{knitrout}
\begin{figure}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}
<<state plots, warning=FALSE, message=FALSE, echo=FALSE>>=
temp5 <- tail(getpars(fitsin5s),10)
temp6 <- tail(getpars(fithomo6s),12)
dat1 <- data.frame(log_mean=temp5[c(1,3,5,7,9)],
                     log_sd=temp5[c(2,4,6,8,10)], states="5") 
dat2 <- data.frame(log_mean=temp6[c(1,3,5,7,9,11)],
                     log_sd=temp6[c(2,4,6,8,10,12)],states="6") 
dat1 <- dat1[order(dat1$log_mean),]
dat2 <- dat2[order(dat2$log_mean),]

dat <- rbind(dat1,dat2)
ggplot(dat,aes(log_mean,log_sd,color=states))+
    geom_point()+geom_line() +theme_bw() +
    ggtitle("State Parameters")

@
\caption{States}
\end{figure}
\end{knitrout}

@@

>>>>

\documentclass{article}
\usepackage{natbib}
\usepackage{amsmath}
\usepackage{alltt}
\usepackage[utf8]{inputenc} % for accented characters
%% stuff for editing
%\usepackage[markup=nocolor,addedmarkup=bf,deletedmarkup=sout]{changes}
%% to suppress notes & comments: \usepackage[final]{changes}
\usepackage[backgroundcolor=lightgray,textsize=tiny]{todonotes}
\usepackage{setspace}
\bibliographystyle{chicago}
\title{Incorporating periodic variability in hidden Markov models for animal movement}
\author{Michael Li and Benjamin M. Bolker }
\date{\today}

@@

\doublespacing
<<pkgs,message=FALSE,warning=FALSE,echo=FALSE>>=
library("ggplot2"); theme_set(theme_bw())
scale_colour_discrete <- function(...,palette="Set1")
    scale_colour_brewer(...,palette=palette)
scale_fill_discrete <- function(...,palette="Set1")
    scale_fill_brewer(...,palette=palette)
zmargin <- theme(panel.margin=grid::unit(0,"lines"))
library("knitr")
opts_chunk$set(fig.width=6,fig.height=4)
library("depmixS4")
library("plyr")
library("dplyr")
library("reshape2")
@

@@

Our primary goal in this paper is to introduce the use of
HMMs with temporally varying transition probabilities -- in particular,
transition probabilities that follow a diurnal cycle -- for modeling
animal movement recorded over long time scales.  
In addition to simulation-based examples,
we also re-analyze data from 
\cite{kerk2015hidden}, who used temporally homogeneous hidden semi-Markov
models (HSMMs: an extension of HMMs that allow
flexible modelling of the distribution of \emph{dwell times}, the 
lengths of consecutive occupancy of a behavioural state) to
describe the movement and putative underlying
behavioural states of Florida panthers (\emph{Puma concolor coryi}).

@@

GPS collars were fitted to 18 Florida panthers in 2005-2012 by Florida
Fish and Wildlife and Conservation Commission staff using
trained hounds and houndsmen. Of these animals, 13 had 
sufficient data to be used by \cite{kerk2015hidden}.
\todo{not sure quite where we should use {\tt citeauthor} vs. {\tt cite} \ldots}
Here we focus
on the four cats with the most data (all with approximately 10,000-15,000 observations: see Table~1 in Supplementary Material), 
in part because our goal is to understand
the issues that arise when simple models are fitted to large data sets, and 
in part because the general trend in telemetry studies is toward larger data sets.  As is typical in studies of animal movement, 
we took first differences of the data by decomposing
contiguous sequences of hourly GPS coordinates into 
successive step lengths (in meters) and turning angles (in radians)
\citep{turchin1998quantitative,kerk2015hidden}.

\cite{kerk2015hidden} used hidden semi-Markov models (HSMM), 
an extension of HMM that permits explicit modelling of dwell times 
\citep{langrock_flexible_2012}, considering both Poisson and
negative binomial distributions for dwell times.  As shown 
by \cite{kerk2015hidden} (Figure S3b, top row, middle panel),
the estimated shape parameter of the negative binomial dwell time
distribution was typically close to 1 ($\approx 0.4-1.6$; confidence
intervals were not calculated), implying that a geometric distribution
(i.e., negative binomial with shape=1) might be adequate. In turn,
this suggests that we might not lose much accuracy by reverting to 
a simpler HMM framework, which makes precisely this assumption.
\todo{could say here (or maybe better in the conclusions) that
revisiting HSMM in this context would be worthwhile future work. 
ML: conclusion is good, I still need to add viterbi paragraph}

\cite{kerk2015hidden} considered time-homogeneous models with 
a variety of candidate distributions %
--- log-Normal, Gamma, and Weibull distributions for step lengths
and von Mises and wrapped Cauchy distributions for the turning angle --- %
concluding on the basis of the Akaike information criterion (AIC) 
that Weibull step length and wrapped Cauchy turning angle distributions
were best.  
Since our analysis aims for simplicity and qualitative
conclusions rather than for picking the very best
predictive model, we focus on models that treat each step as
a univariate, log-Normally distributed observation, glossing
over both the differences in shape between
the three candidate step-length distributions and the effects
of considering multivariate (step length plus turning angle)
observations.  However, we do briefly compare log-Normal and Weibull
step-length distributions\todo{have we run Gamma as well?}, with and without 
a von Mises-distributed turning angle included in the model (Figure~\ref{fig:BICred_plot}).\todo{W/VM/HMM (blue/solid/left panel) should correspond to vdK's results, but that has $\dbic=0$ for 5 states rather than 6. Is that right? Is that HMM/HSMM diff?}
(Note that most movement analyses, including
\cite{kerk2015hidden}, are only partially multivariate, 
treating step length and turning angle at a particular time
as independent observations while neglecting possible correlations
between them.)

\cite{kerk2015hidden} used the Bayesian (Schwarz) information criterion (BIC)
to test the relative penalized goodness of fit for
models ranging from 2 to 6 latent states.
In general, BIC values decreased as the number of states
increased from 
3 to 6 states,
suggesting that the 6-state model was
favoured statistically; however,
the authors used 3-state models in most
of their analyses for ease of biological interpretation.
We follow \cite{kerk2015hidden} in using BIC-optimality (i.e., minimum
BIC across a family of models) as the criterion for identifying
the best model, because we are interested in explaining the 
data generation process by identifying the ``true'' number 
of underlying movement states.  Using BIC also simplifies 
evaluation of model selection procedures; 
it is easier to test whether our model selection
procedure has selected the model used to simulate the data,
rather than testing whether it has selected the model with
the minimal Kullback-Leibler distance
\citep{Richards2005}. We recognize that 
ecologists will often be interested in maximizing predictive
accuracy rather than selecting a true model, and that as
usual in ecological systems the true model will be far more
complicated than any candidate model \citep{BurnhamAnderson1998};
we suspect that the qualitative conclusions stated here
for BIC-optimality will carry over to analyses using AIC instead.

@@

Eq.~\ref{eq:HMMlik} gives the likelihood of the observed sequence 
given (conditional on) a particular
hidden sequence. 
In order to calculate the overall, unconditional (or marginal) 
likelihood of the 
observed sequence, we need to average over all possible hidden sequences. 
There are several efficient algorithms for computing the marginal likelihood and
numerically estimating parameters \citep{zucchini_hidden_2009};
we use those implemented in the \texttt{depmixS4} package for R
\citep{visser2010depmixs4,R}.

For any $n$-state HMM, we need to define a $n \times n$ matrix that specifies the probabilities $\pi_{ij}$ of being in movement states $j$ at time $t+1$ given that the individual is in state $i$.  The FMM is a special case of HMM where the probabilities of \emph{entering} a given state are identical across all states --- i.e., the probability of occupying a state at the next time step is independent of the current state occupancy. It can be modelled in the HMM framework by setting the transition probabilities  $\pi_{ij} = \pi_{i*}$. \todo{is this right?}

In any case, the transition matrix $\pi_{ij}$ must respect the constraints that (1) all probabilities are between 0 and 1 and (2) transition probabilities out of a given state sum to 1.
As is standard for HMMs with covariates \citep{visser2010depmixs4}, we define this multinomial logistic model in terms of a linear predictor $\eta_{ij}$, where $\eta_{i1}$ is set to 1 without loss of generality (i.e. we have only $n \times (n-1)$ distinct parameters; we nevertheless index $j$ from 2 to $n$ for notational clarity):
\begin{equation}
\begin{split}
\pi_{ij} & = \exp(\eta_{ij}(t))/\left(1+\sum_{j=2}^{n}\exp(\eta_{ij}(t))\right), \textrm{ for } j={2,...,n} \\
\pi_{i1} & = 1 - \sum_{j=2}^{n}\pi_{ij}
\end{split}
\end{equation}

We considered three different transition models for diurnal variation in behaviour, incorporating hour-of-day as a covariate following the general approach of \cite{morales_extracting_2004} of allowing for a covariate-dependent transition matrix.

\begin{description}
\item[Multiple block transition] Here we assume piecewise-constant transition probabilities. The transition probability $\pi_{ij}$ is a function of time (hour of day), where it is assigned to one of $M$ different time blocks:
$$
\eta_{ij}(t) = \sum_{m=1}^{M}a_{ijm} \delta_{m=t}
$$
where $a_{ijm}$ are parameters, and $\delta_{m=t}$ is a Kronecker delta or  indicator variable (1 for the time block at the corresponding time $t$, and 0 otherwise).  
\item[Quadratic transition model] We assume the elements of the linear predictor
are quadratic functions of hour. Note that the quadratic model is not diurnally continuous, i.e. there is no constraint forcing $\eta_{ij}(0)=\eta_{ij}(1)$; imposing a diurnal continuity constraint would collapse the model to a constant.
$$
\eta_{ij}(t) = b_{ij1}+b_{ij2}(\frac{t}{24})+b_{ij3}(\frac{t}{24})^{2}
$$
\item[Sinusoidal transition model] A sinusoidal model with a period of 24 hours is identical in complexity to the quadratic model, but automatically satisfies the diurnal continuity constraint.
$$
\eta_{ij}(t)= b_{ij1}+b_{ij2}\cos\left(\frac{2\pi t}{24}\right)+b_{ij3}\sin\left(\frac{2\pi t}{24}\right)
$$
\end{description}

Model complexity and number of parameters increase as the number of latent states increase, FMM to HMM, and lastly, FMM and HMM incorportating temporal heterogeneity.

\begin{table}[]
\centering
\caption{Number of Parameters for each model type}
\label{Cat-ID}
\begin{tabular}{cccc}
 Type & Model  & Number of States & Number of Parameters  \\
  FMM & FMM & 3 & 9    \\
  FMM & FMM & 4 & 12    \\
  FMM & FMM & 5 & 15   \\
  FMM & FMM & 6 & 18  \\
  FMM + TH & FMM + TH (Sin)  & 3 & 15    \\
  FMM + TH & FMM + TH (Sin)  & 4 & 20    \\
  FMM + TH & FMM + TH (Sin)  & 5 & 25    \\
  FMM + TH & FMM + TH (Sin)  & 6 & 30    \\
  HMM & HMM & 3 & 18    \\
  HMM & HMM & 4 & 28    \\
  HMM & HMM & 5 & 40   \\
  HMM & HMM & 6 & 54  \\
  HMM + TH & HMM + TH (Block)  & 3 & 36    \\
  HMM + TH & HMM + TH (Block)  & 4 & 60    \\
  HMM + TH & HMM + TH (Block)  & 5 & 90    \\
  HMM + TH & HMM + TH (Block)  & 6 & 126    \\
  HMM + TH & HMM + TH (Quad)  & 3 & 36    \\
  HMM + TH & HMM + TH (Quad)  & 4 & 60    \\
  HMM + TH & HMM + TH (Quad)  & 5 & 90    \\
  HMM + TH & HMM + TH (Quad)  & 6 & 126    \\
  HMM + TH & HMM + TH (Sin)  & 3 & 36    \\
  HMM + TH & HMM + TH (Sin)  & 4 & 60    \\
  HMM + TH & HMM + TH (Sin)  & 5 & 90    \\
  HMM + TH & HMM + TH (Sin)  & 6 & 126    \\
  HMM + TH & HMM + TH (Hourly)  & 3 & 225    \\
  HMM + TH & HMM + TH (Hourly)  & 4 & 396    
\end{tabular}
\end{table}

\todo{Discuss model complexity/parameter-number computations here, or with each model.  Also discuss the ``hourly'' model here.  Make the table into a figure!}

@@

<<BICred_plot,warning=FALSE,echo=FALSE,message=FALSE,fig.cap="Change in BIC across models, cat \\#130.  $\\dbic$ is scaled individually per model, so that $\\dbic=0$ represents the BIC-optimal number of states for a particular model.",fig.width=6,fig.height=4>>=
load("cat1allmod.RData")
df2 <- transform(allmodsumdf,
            distrib=ifelse(grepl("LN",model),"log-Normal","Weibull"),
            turn_angle=ifelse(grepl("VM",model),"yes","no"),
            homog=ifelse(grepl("HMM",model),"homogeneous HMM","time-heterogeneous HMM"))
ggplot(df2, aes(x=nstates,y=deltaBIC,colour=distrib,lty=turn_angle))+
    geom_line()+
    facet_wrap(~ homog)+
    scale_colour_brewer(name="step length\ndistribution",palette="Set1")+
    ## scale_size_continuous( name="# parameters") +
    #    geom_point(aes(size=parameters))+
    scale_linetype(name="von Mises\nturning angle?")+
    labs(y=expression(Delta*"BIC"),
         x="Number of States")+
    geom_hline(yintercept=0,colour="gray")+
    zmargin
@

@@

<<adj_BIC_comparisons, warning=FALSE, echo=FALSE, message=FALSE,fig.cap="Overall Adjusted BIC Comparisons between different HMM models">>=
load("cat1sumdat.RData")
ggplot(sumdat, aes(x=nstates,y=deltaBIC,colour=model))+facet_wrap(~ type,ncol=4)+
    scale_size_continuous( name="Num of Parameters") +
    geom_point(aes(size=parameters))+
    scale_colour_brewer(palette="Dark2")+
    labs(y=expression(paste(Delta, "BIC")),
         x="Number of States") +
    geom_line() + zmargin
@

@@

<<avg_step_length_by_time, message=FALSE, echo=FALSE, warning=FALSE, fig.cap="Average step length by time of day">>=
  simdf <- simdat %>% dplyr:::select(-obs) %>% group_by(time) %>% summarise_each(funs(mean))
  obsdf <- simdat %>% dplyr:::select(c(obs,time)) %>% filter(!is.na(obs)) %>% group_by(time) %>% summarise_each(funs(mean)) %>% dplyr:::select(obs)
  avgdf <- cbind(simdf,obsdf)
  avgdf2 <- melt(avgdf,'time')
ggplot(avgdf2,aes(x=time,y=value,colour=variable)) + 
    geom_point() +geom_line()+ xlab("Time of Day") + ylab("Step lengths (log10)")+
    scale_x_continuous(breaks=0:23)
@

@@

<<acf_plot, message=FALSE, echo=FALSE, warning=FALSE, fig.cap="ACF">>=
acffun <- function(sim){
  a <- acf(sim,na.action=na.pass,plot=FALSE)
  df <- data.frame(ACF=c(a$acf),lag=a$lag)
  return(df)
}

  temp <- simdat %>% dplyr:::select(-time)%>%ldply(.,acffun)
  ggplot(temp,aes(lag,ACF,colour=.id))+
    geom_point()+geom_line()
@

@@

<<state_plots, warning=FALSE, message=FALSE, echo=FALSE, fig.cap="States">>=
temp5 <- tail(getpars(fitsin5s),10)
temp6 <- tail(getpars(fithomo6s),12)
dat1 <- data.frame(log_mean=temp5[c(1,3,5,7,9)],
                     log_sd=temp5[c(2,4,6,8,10)], states="5") 
dat2 <- data.frame(log_mean=temp6[c(1,3,5,7,9,11)],
                     log_sd=temp6[c(2,4,6,8,10,12)],states="6") 
dat1 <- dat1[order(dat1$log_mean),]
dat2 <- dat2[order(dat2$log_mean),]

dat <- rbind(dat1,dat2)
ggplot(dat,aes(log_mean,log_sd,color=states))+
    geom_point()+geom_line()
@

@@

\appendix
\section{Supplementary material}
\begin{table}[]
\centering
\caption{Cat-ID}
\label{Cat-ID}
\begin{tabular}{ccc}
  (van de Kerk 2015) & (IR@UF)  & Number of Obserations  \\
  130 & 1   & 10286    \\
  131 & 2   & 9458    \\
  48  & 14  & 14645    \\
  94  & 15  & 10250  
\end{tabular}
\end{table}
